Designing Qusecs Optimally: Quasi-uniform or Self-similar Layered Materials as Recursively Decomposed Solutions of Geometric Constraint Systems

### 2014-12-18 doc/GMP2015/s04_recombination.tex

The main barrier in recombination when given an optimal DR-plan (of
smallest possible size or max fan-in),  is that the number of children
(resp. number of roots) - and correspondingly the  size and complexity
of the (indecomposable) algebraic system $I$ to be solved - could be
arbitrarily large as a function of the size of the input system. This
difficulty can persist even after optimal parametrization of the
indecomposable system $I$ as in \uncited to minimize its algebraic
complexity.

@@

It is immediately clear that indecomposable graphs $G$ that belong in
OMD$_k$ for small $k$ and $s$  lend themselves to modification  into
decomposable graphs satisfying Criteria (a) and (b) above. However, it
is not clear how Criterion (c) is met by OMD graphs.
%
\subsection{Using Convex Cayley Configuration Spaces}
\label{sec:2-tree-reduction}
%
Next we provide the necessary background to describe a specific
approach for achieving the requirements (a)-(c) mentioned above, by
restricting the class of reduced graphs $G' = G\setminus E_1$ and
their isostatic completions $H$ in the above definition of the OMD
problem, and using a key theorem of Convex Cayley configuration spaces
\uncited. This theorem characterizes the class of graphs $H$ and non-
edges $F$ (Cayley parameters), such that the set of vectors
$\lambda_F$ of  attainable lengths of the nonedges $F$ is always
convex for any given lengths $\delta$ for the edges of $H$ (i.e, over
all the realization s of the bar-joint constraint system or linkage
$(H,\delta)$ in 2 dimensions). This set is called the (2-dimensional)
\dfn{Cayley configuration space} of the linkage $(H,\delta)$ on the
Cayley parameters $F$, denoted $\Phi_F(H,\delta)$ and can be viewed as
a ``projection'' of the cartesian realization space of $(H,\delta)$ on
the Cayley parameters $F$. Such graphs $H$ are said to have
\dfn{convexifiable Cayley configuration spaces with parameters $F$}
(short: \dfn{convexifiable}). To state the theorem, we first have to
define the notion of \dfn{2-sums} and \dfn{2-trees}. Let $H_1$ and
$H_2$ be two graphs on disjoint sets of vertices $V_1$ and $V_2$, with
edge sets $E_1$ and $E_2$ containing edges $(u,v)$ and $(w,x)$
respectively.  A \dfn{2-sum} of $H_1$ and $H_2$ is a graph $H$
obtained by taking the union of $H_1$ and $H_2$ and identifying $u=w$
and $v=w$. In this case, $H_1$ and $H_2$ are called \dfn{2-sum
components} of $H$. A \dfn{minimal 2-sum component} of $H$ is  one
that cannot be further split into 2-sum components. A \dfn{2-tree} is
recursively obtained by taking a 2-sum of two 2-trees, with the base
case of 2-tree being a triangle. A \dfn{partial 2-tree} is a 2-tree
minus some edges. Partial 2-trees are characterized as having $K_4$ as
forbidden minor are also called series parallel graphs.

\begin{theorem}\label{theorem:convexcayley}
    \uncited $H$ has a convexifiable Cayley configuration space  with
    parameters $F$ if and only if for each $f\in F$  all the minimal
    2-sum components of $H\cup F$ that contain both endpoints of $f$
    are partial 2-trees. The Cayley configuration space
    $\Phi_F(H,\delta)$ of a bar-joint system or linkage $(H,\delta)$
    is a convex polytope. When $H\cup F$ is a 2-tree, the bounding
    hyperplanes of this polytope are triangle inequalities relating
    the lengths of edges of the triangles in $H\cup F$.
\end{theorem}

@@

\noindent
\note A huge advantage of the convex Cayley method is that it is
completely unaffected when $\delta$ are intervals rather than exact
values~\cite{GaoSitharam}.
%cite my paper with Heping!!!

\medskip\noindent
\textbf{Example}. \todo{This uses different example format, not their environment. Should this, or section 3 change?}
A graph $G=K_{3,3}$  cannot be decomposed into any nontrivial
isostatic graphs, i.e, its DR-plan has a root and 9 children
corresponding to the 9 edges. Solving or recombining the system
$(G,\delta)$ corresponding to the root of this DR-plan involves
solving a quadratic system with 8 equations and variables. Instead of
simultaneously solving this system, we could instead use the fact that
$G=K_{3,3}$ is in OMD$_1$: remove the edge $e$ in Figure
\ref{fig:omd_k33_example} to give a partial 2-tree $G'$. Now add the
nonedge $f$ to give a 2-tree $H$ with a DR-plan of size 2. The Cayley
configuration space $\Phi_f(G', \delta_{E\setminus e})$ is a single
interval of attainable length values $\lambda_F$ for the edge $f$.
When $\delta$ is generic, i.e, does not admit collinearities or
coincidences in the realizations of $(G,\delta)$, the realization
space of $(H, \left<\delta_{E\setminus e}, \lambda_f\right>)$ has 16
solutions $q^p_{\lambda_f}$ (modulo orientation preserving
isometries), with distinct orientation types $\sigma_p$ (two
orientation choices for each of the 4 triangles) that can be obtained
by solving a sequence of 4 single quadratics in 1 variable (DR-plan of
size 2). By subdivided binary search in the interval $\lambda_f \in
\Phi_f(G', \delta_{E\setminus e})$, the desired solution $p$  of
$(G,\delta)$ is found when the length of the nonedge $e$ in  the
realization $q^p_{\lambda_f}$ is $\delta_e$.
%
\subsection{Optimized Modification by Enlarging the class of Reduced
Graphs}
\label{sec:tdecomp}
%
It is possible in principle to decrease $k$ for a OMD$_k$ graph (i.e,
the number of edges to be removed to ensure an isostatic completion
that is decomposable with a small DR-plan) by considering reduced
graphs $G'$ (and modified graphs $H$) that come from a larger class
than partial 2-trees but nevertheless have convex Cayley configuration
spaces at least when the realization space is restricted to a
sufficiently comprehensive orientation type. In particular, the so-
called \dfn{tree-decomposable graphs of low Cayley complexity}
\uncited include the partial 2-trees and many others that are not
partial 2-trees. These too result in DR-plans of size 2 or 3, putting
$G$ in the class OMD$_k$ and thus meeting Criteria (a) and (b). The
Criterion (c) is met---for example when $k=1$---because 1-dof Cayley
configuration spaces of linkages based on such graphs $G'$ are known
to be single intervals when a comprehensive orientation type
$\sigma_p$ of the sought solution $p$ is given. In addition, the
bounds of these intervals are of low algebraic complexity. More
precisely, the bounds  can themselves be computed using a DR-plan of
size 2 or 3, i.e, the computation of these bounds is tree-
decomposable. Alternatively, the bounds are in a simple quadratic or
radically solvable extension field over the rationals, or can be
computed by solving a triangularized system of quadratics.

>>>>

The main barrier in recombination when given an optimal DR-plan (of
smallest possible size or max fan-in),  is that the number of children
(resp. number of roots) - and correspondingly the  size and complexity
of the (indecomposable) algebraic system $I$ to be solved - could be
arbitrarily large as a function of the size of the input system. This
difficulty can persist even after optimal parametrization of the
indecomposable system $I$ as in \cite{sitharam2010optimized} to minimize its algebraic
complexity.

@@

It is immediately clear that indecomposable graphs $G$ that belong in
OMD$_k$ for small $k$ and $s$  lend themselves to modification  into
decomposable graphs satisfying Criteria (a) and (b) above. However, it
is not clear how Criterion (c) is met by OMD graphs.
%
\subsection{Using Convex Cayley Configuration Spaces}
\label{sec:2-tree-reduction}
%
Next we provide the necessary background to describe a specific
approach for achieving the requirements (a)-(c) mentioned above, by
restricting the class of reduced graphs $G' = G\setminus E_1$ and
their isostatic completions $H$ in the above definition of the OMD
problem, and using a key theorem of Convex Cayley configuration spaces
\cite{sitharam2010convex}. This theorem characterizes the class of graphs $H$ and non-
edges $F$ (Cayley parameters), such that the set of vectors
$\lambda_F$ of  attainable lengths of the nonedges $F$ is always
convex for any given lengths $\delta$ for the edges of $H$ (i.e, over
all the realization s of the bar-joint constraint system or linkage
$(H,\delta)$ in 2 dimensions). This set is called the (2-dimensional)
\dfn{Cayley configuration space} of the linkage $(H,\delta)$ on the
Cayley parameters $F$, denoted $\Phi_F(H,\delta)$ and can be viewed as
a ``projection'' of the cartesian realization space of $(H,\delta)$ on
the Cayley parameters $F$. Such graphs $H$ are said to have
\dfn{convexifiable Cayley configuration spaces with parameters $F$}
(short: \dfn{convexifiable}). To state the theorem, we first have to
define the notion of \dfn{2-sums} and \dfn{2-trees}. Let $H_1$ and
$H_2$ be two graphs on disjoint sets of vertices $V_1$ and $V_2$, with
edge sets $E_1$ and $E_2$ containing edges $(u,v)$ and $(w,x)$
respectively.  A \dfn{2-sum} of $H_1$ and $H_2$ is a graph $H$
obtained by taking the union of $H_1$ and $H_2$ and identifying $u=w$
and $v=w$. In this case, $H_1$ and $H_2$ are called \dfn{2-sum
components} of $H$. A \dfn{minimal 2-sum component} of $H$ is  one
that cannot be further split into 2-sum components. A \dfn{2-tree} is
recursively obtained by taking a 2-sum of two 2-trees, with the base
case of 2-tree being a triangle. A \dfn{partial 2-tree} is a 2-tree
minus some edges. Partial 2-trees are characterized as having $K_4$ as
forbidden minor are also called series parallel graphs.

\begin{theorem}\label{theorem:convexcayley}
    \cite{sitharam2010convex} $H$ has a convexifiable Cayley configuration space  with
    parameters $F$ if and only if for each $f\in F$  all the minimal
    2-sum components of $H\cup F$ that contain both endpoints of $f$
    are partial 2-trees. The Cayley configuration space
    $\Phi_F(H,\delta)$ of a bar-joint system or linkage $(H,\delta)$
    is a convex polytope. When $H\cup F$ is a 2-tree, the bounding
    hyperplanes of this polytope are triangle inequalities relating
    the lengths of edges of the triangles in $H\cup F$.
\end{theorem}

@@

\noindent
\note A huge advantage of the convex Cayley method is that it is
completely unaffected when $\delta$ are intervals rather than exact
values~\cite{sitharam2010convex}.

\medskip\noindent
\textbf{Example}. \todo{This uses different example format, not their environment. Should this, or section 3 change?}
A graph $G=K_{3,3}$  cannot be decomposed into any nontrivial
isostatic graphs, i.e, its DR-plan has a root and 9 children
corresponding to the 9 edges. Solving or recombining the system
$(G,\delta)$ corresponding to the root of this DR-plan involves
solving a quadratic system with 8 equations and variables. Instead of
simultaneously solving this system, we could instead use the fact that
$G=K_{3,3}$ is in OMD$_1$: remove the edge $e$ in Figure
\ref{fig:omd_k33_example} to give a partial 2-tree $G'$. Now add the
nonedge $f$ to give a 2-tree $H$ with a DR-plan of size 2. The Cayley
configuration space $\Phi_f(G', \delta_{E\setminus e})$ is a single
interval of attainable length values $\lambda_F$ for the edge $f$.
When $\delta$ is generic, i.e, does not admit collinearities or
coincidences in the realizations of $(G,\delta)$, the realization
space of $(H, \left<\delta_{E\setminus e}, \lambda_f\right>)$ has 16
solutions $q^p_{\lambda_f}$ (modulo orientation preserving
isometries), with distinct orientation types $\sigma_p$ (two
orientation choices for each of the 4 triangles) that can be obtained
by solving a sequence of 4 single quadratics in 1 variable (DR-plan of
size 2). By subdivided binary search in the interval $\lambda_f \in
\Phi_f(G', \delta_{E\setminus e})$, the desired solution $p$  of
$(G,\delta)$ is found when the length of the nonedge $e$ in  the
realization $q^p_{\lambda_f}$ is $\delta_e$.
%
\subsection{Optimized Modification by Enlarging the class of Reduced
Graphs}
\label{sec:tdecomp}
%
It is possible in principle to decrease $k$ for a OMD$_k$ graph (i.e,
the number of edges to be removed to ensure an isostatic completion
that is decomposable with a small DR-plan) by considering reduced
graphs $G'$ (and modified graphs $H$) that come from a larger class
than partial 2-trees but nevertheless have convex Cayley configuration
spaces at least when the realization space is restricted to a
sufficiently comprehensive orientation type. In particular, the so-
called \dfn{tree-decomposable graphs of low Cayley complexity}
\cite{sitharam2011cayleyI,sitharam2011cayleyII} include the partial 2-trees and many others that are not
partial 2-trees. These too result in DR-plans of size 2 or 3, putting
$G$ in the class OMD$_k$ and thus meeting Criteria (a) and (b). The
Criterion (c) is met---for example when $k=1$---because 1-dof Cayley
configuration spaces of linkages based on such graphs $G'$ are known
to be single intervals when a comprehensive orientation type
$\sigma_p$ of the sought solution $p$ is given. In addition, the
bounds of these intervals are of low algebraic complexity. More
precisely, the bounds  can themselves be computed using a DR-plan of
size 2 or 3, i.e, the computation of these bounds is tree-
decomposable. Alternatively, the bounds are in a simple quadratic or
radically solvable extension field over the rationals, or can be
computed by solving a triangularized system of quadratics.

### 2014-12-18 doc/GMP2015/s08_pinnedline.tex
White space fixes.

\dfn{Collagen} is an important protein material in biological tissues with highly elastic mechanical property.
\dfn{Cellulose} is the most important constituent of the cell wall of plants (see Figure~\ref{fig:cellulose}).
Both of these substances consist of a large number \dfn{microfibrils},
%fibril itself straight?
%Each fibril is attached to some fixed larger organelle/membrane at one spot (typically on the boundary).
%It is additionally
each of which is cross-linked at 2 places with some two other fibrils,
where the \dfn{cross-linking} is like an incidence constraint that the
two crosslinked fibrils can slide with each other  \uncited (see
Figure~\ref{fig:crosslink}).

@@

\begin{subfigure}{.49\linewidth}
  \centering
  \includegraphics[width=.8\linewidth]{img/AFM_Innventia_nanocellulose}
  \caption{}
  \label{fig:cellulose}
\end{subfigure}
\begin{subfigure}{.49\linewidth}
  \centering
  \includegraphics[width=\linewidth]{img/crosslink}
  \caption{}
  \label{fig:crosslink}
\end{subfigure}%
\caption{(a) Microfibrils of carboxymethylated nanocellulose adsorbed on a silica surface \uncited. (b) Cross-linking of cellulose microfibrils \uncited.}
\end{figure}

@@

A \dfn{pinned line-incidence system} $(G,\delta)$ is a graph $G=(V,E)$
together with parameters $\delta$ specifying $|E|$ \textbf{pins} with
fixed positions in $\mathbb{R}^2$, such that each edge is constrained
to lie on a line passing through the corresponding pin, i.e.\ $\delta:
E \rightarrow \mathbb{R}^2$.
%
% $G=(V,E, w_V, w_E)$
%%is a \textbf{2-dimensional pinned line-incidence graph},
% has weight functions $w_V(v) = 2, w_E(e) = 1$ for all $v \in V, e \in E$.
%Additionally, there is  a  set $X$ of $|E|$ \textbf{pins} with fixed positions in $\mathbb{R}^2$,
%such that  %every edge $e \in E$ corresponds to a pin $x \in X$ where
%for each edge $e \in E$, there is a uniquely corresponding pin $x \in X$,
%such that $e$ is constrained to lie on a line passing through $x$.
%%There exists a set $X$ of $|E|$ pins fixed in $\mathbb{R}^2$, such that each edge $e_i$
%
A pinned line-incidence graph $G$ is rigid if $|E| = 2|V|$ and $|E'| \le 2|V'|$ for every induced subgraph $(V',E')$ \uncited. Note that no trivial motion exists since the pins have fixed positions on the plane.
In particular, both a single vertex and a single edge are under-constrained graphs.
%and the smallest over-constrained graph has $6$ vertices.

>>>>

\dfn{Collagen} is an important protein material in biological tissues with highly elastic mechanical property \cite{buehler2008nanomechanics}.
\dfn{Cellulose} is the most important constituent of the cell wall of plants (see Figure~\ref{fig:cellulose}) \cite{fall2013physical,smith1971plant}.
Both of these substances consist of a large number \dfn{microfibrils},
%fibril itself straight?
%Each fibril is attached to some fixed larger organelle/membrane at one spot (typically on the boundary).
%It is additionally
each of which is cross-linked at 2 places with some two other fibrils,
where the \dfn{cross-linking} is like an incidence constraint that the
two crosslinked fibrils can slide with each other (see
Figure~\ref{fig:crosslink}).

@@

\begin{subfigure}{.49\linewidth}
  \centering
  \includegraphics[width=.8\linewidth]{img/AFM_Innventia_nanocellulose}
  \caption{}
  \label{fig:cellulose}
\end{subfigure}
\begin{subfigure}{.49\linewidth}
  \centering
  \includegraphics[width=\linewidth]{img/crosslink}
  \caption{}
  \label{fig:crosslink}
\end{subfigure}%
\caption{(a) Microfibrils of carboxymethylated nanocellulose adsorbed on a silica surface \cite{wikimediacommons2010afm}. (b) Cross-linking of cellulose microfibrils \cite{wikimediacommons2007plant}.}
\end{figure}

@@

A \dfn{pinned line-incidence system} $(G,\delta)$ is a graph $G=(V,E)$
together with parameters $\delta$ specifying $|E|$ \textbf{pins} with
fixed positions in $\mathbb{R}^2$, such that each edge is constrained
to lie on a line passing through the corresponding pin, i.e.\ $\delta:
E \rightarrow \mathbb{R}^2$.
%
% $G=(V,E, w_V, w_E)$
%%is a \textbf{2-dimensional pinned line-incidence graph},
% has weight functions $w_V(v) = 2, w_E(e) = 1$ for all $v \in V, e \in E$.
%Additionally, there is  a  set $X$ of $|E|$ \textbf{pins} with fixed positions in $\mathbb{R}^2$,
%such that  %every edge $e \in E$ corresponds to a pin $x \in X$ where
%for each edge $e \in E$, there is a uniquely corresponding pin $x \in X$,
%such that $e$ is constrained to lie on a line passing through $x$.
%%There exists a set $X$ of $|E|$ pins fixed in $\mathbb{R}^2$, such that each edge $e_i$
%
A pinned line-incidence graph $G$ is rigid if $|E| = 2|V|$ and $|E'| \le 2|V'|$ for every induced subgraph $(V',E')$ \cite{sitharam2014incidence}. Note that no trivial motion exists since the pins have fixed positions on the plane.
In particular, both a single vertex and a single edge are under-constrained graphs.
